{\plant}: A package for modelling forest trait ecology \& evolution

### 2015-11-09 ms.tex
Incorporate feedback from Ulf and Ake
Comments provided via email, have implemented some but not all of suggested changes

Changes reviwed with Mark Westoby. He suggested also removing quotes on some
items.

\title{{\plant}: A package for modelling forest trait ecology \& evolution}
\author{Daniel S. Falster\textdagger\textasteriskcentered, Richard G. FitzJohn\textdagger, and Mark Westoby}
\affiliation{
Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
\textdagger These authors contributed equally.\\
\textasteriskcentered Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
A manuscript in consideration as an Applications Note for
publication in MEE as part of the Special Feature \emph{Demography
  beyond the Population}.\\
Word count: ~4200 words}
\date{}

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured:
  larger plants shade smaller plants while also expending
  proportionately more energy on building and maintaining woody stems.
  Although the importance of size-
  structure for demography is widely recognised, many models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the {\plant} package, an
  extensible framework for modelling size- and trait--based demography,
  ecology and evolution in simulated forests.
  At its core, {\plant} is an
  individual-based model where plant physiology and demography are mediated by
  traits. Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants can be
  resolved either in a stochastic, finite-sized patch; or deterministically, using
  novel extensions of the ``Escalator Boxcar Train''
  technique. These dynamics can be integrated into metapopulation--level estimates of
  invasion fitness and vegetation structure. Because fitness emerges as a
  function of traits, plant
  provides a novel arena for exploring eco-evolutionary dynamics.
\item
  {\plant} is an open source R package and is available at
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
  While accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation to capture trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
\item
  {\plant} provides a transparent platform for investigating how
  physiological rules and functional trade-offs interact with competition and
  disturbance regimes to influence vegetation demography, structure and
  diversity.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}.
%
As an individual increases its leaf area, it increases its potential
to generate photosynthate.
%
On the other hand as individuals grow larger they must allocate increasing
fractions of their photosynthetic income to activities other
than building new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
Ontogenetic
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates
\citep{Wright-2010}, while seed size and height
at maturation determine the start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Shugart-1980, Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size-structure for trait evolution, vegetation
dynamics and diversity has been long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size-structure
entirely, or only include coarse approximations.
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
  Tilman-1985, Geritz-1998, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes are influenced by
size-related traits such as adult and seed size, but detailed
size-structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Similarly, many of the models used to study global vegetation dynamics across
the last 20 years have not included size-structured demography
\citep[for comparisons of some major models see][]{Sitch-2008,
Dekauwe-2014}.
%
While the importance of explicitly including size-structured dynamics is
increasingly recognised in both areas of evolutionary ecology
\citep[e.g.][]{Falster-2015} and vegetation dynamics
\citep[e.g.][]{Moorcroft-2001, Purves-2008, Smith-2014,
Weng-2015, Sakschewski-2015}, our understanding on how these features
impact on the ecology and evolution of vegetation communities remains
relatively limited.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of metapopulations structured by a prevailing
disturbance regime.
%
Our own purpose in developing {\plant} has been mainly to investigate how species
differing in traits may be able to coexist with one another
\citep[following ][]{Falster-2011, Falster-2015}; as size-structured models
provide unique opportunities for studying coexistence via differentiation in
successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
At the same time, we expect {\plant} to be useful for modelling other demographic phenomena.

Broadly, the {\plant} package falls into a class we refer to as Trait- Size- and
Patch- structured Models \citep[\textsc{TSPMs}, following][]{Falster-2011}. \textsc{TSPMs} are direct
descendants from the ``gap'' models developed in 1980s
\citep[e.g.][]{Shugart-1980, Huston-1987, Kohyama-1993}; other modern examples
include \textsc{ED} \citep{Moorcroft-2001}, \textsc{LPJ- GUESS}
\citep{Smith-2014} and \textsc{LPJmL-FIT} \citep{Sakschewski-2015}. Common to
\textsc{TSPMs} is that they explicitly model size-structured competition
among individual plants within a metapopulation of ``patches'', where
individuals can also differ in their traits and thus demographic behaviour.
\textsc {TSPMs} typically ignore the spatial structure of plants within
patches and the spatial structure of patches relative to one another. Instead
they focus computational effort on resolving the dynamics of size-structured
competition for light and successional turnover. Differences among
\textsc{TSPMs} arise from the core physiological models used (describing
how the demography of individual plants responds to resource availability),
and also from the numerical techniques applied to scaling the physiological model up to
estimate emergent behaviours of individuals, patches and vegetation.

Below we describe the general approach of the {\plant} package, then provide
sections outlining potential applications at nested levels of ecological organisation.
More detailed technical documents are provided as supplementary information
(see Appendices; updated versions of these technical documents are
available within the {\plant} package itself). These outline \ref{sec:demography}) the system of
equations being solved when modelling demography of plants, patches and metapopulations; \ref{sec:FFW16}) a description of the core
physiological model used {\plant}, and \ref{sec:code}) a compendium of worked code examples showing how to interact with the {\plant} package from R.

@@

The {\plant} package implements an individual-based model, meaning the dynamics of the
population arise from rules specifying how individuals grow and interact.
Being driven by traits, the model can be extended to potentially very many species. The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how this is influenced by its
traits, size and light environment.  While {\plant} includes a core physiological model, the present paper uses it only as an example, to illustrate how one can take a given physiological model and simulate ecological and evolutionary outcomes. Thus, we do not aim to justify the particular physiological model used, to provide detailed
comparisons to data, or to promote this physiological model over those
in other \textsc{TSPMs}. (These are all topics deserving substantially
more attention than can be afforded here.)
% TODO: I'm a bit concerned that while we know that this is possible,
% it's not that well developed and does require compiling a new
% version of plant.  So it's not that _usefully_ extendable.  On the
% other hand, the population of users who will do this is not very
% large.  It might be possible to allow runtime linkage though if I
% push a little more of the cpp code into headers and do some tweaking
% in RcppR6...
An important feature of {\plant} considered as a software package is that it allows for users to substitute their own physiological models.

On top of the physiological model, {\plant} implements methods for
population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at the three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at these higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
{\plant} offers the capacity to model emergent behaviour of patches with
specified size, and also a deterministic size-independent approximation
where specifying a patch size is not required. Trait evolution and community
assembly can then be modelled using the estimates of invasion fitness
provided by {\plant}.

\section{Effects of size, trait and environment on demography of individual plants}

@@

The default physiological model used in {\plant} largely reflects
that presented in \citet{Falster-2011,Falster-2015}, but with extensions
allowing for diameter growth also to be estimated. We refer to this
default model as \texttt{FFW16}, reflecting the author's initials and
publication year of this paper. The \texttt{FFW16} model estimates rate
of biomass
production for a plant, given its size and the current light
environment. Gross photosynthetic income is calculated from the total
leaf area and the light distribution across the plant's canopy. Costs
of tissue respiration and turnover are then subtracted. The remaining
biomass production is then allocated between growth and reproduction. The key
outputs needed by higher levels of {\plant} are height
growth rate, mortality rate and rate of seed production (Fig. \ref{fig:schematic}).  Additional quantities are also computed, such as total assimilation
and the respiration, turnover and allocation rates for different
tissues, and these intermediates can also be accessed (for further details on the \texttt{FFW16} model see Appendix
\ref{sec:FFW16} and \ref{sec:code}).

While we have implemented a particular physiological model
(\texttt{FFW16}), the package is designed to allow arbitrary
additional physiologies to be introduced. Users are able to modify the
physiological model in two different ways. First, one can vary the
parameters of the \texttt{FFW16} model directly (see Appendix
\ref{sec:FFW16} for list, or Appendix \ref{sec:code} for code
examples). Second, we allow for changes in key parameters to bring
about changes in other parameters (a hyper-parameterisation), enabling
straightforward modelling of trade-offs. For example, by default the
trait \textsc{lma} (leaf mass per unit leaf area) is used to estimate
the rate of leaf turnover ($k_l$), based on an observed scaling
relationship spanning across diverse vegetation and plant functional
types \citep{Wright-2004}:
$$k_l = \beta_{kl1} \, \left(\frac{\phi}{\phi_0}\right)^{-\beta_{kl2}},$$
where $k_l$ and $\phi$ (lma) are both parameters of the core physiological
model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such
linkages are defined within a user-supplied hyper-parameterisation function
(see Appendix \ref{sec:FFW16} for details, or Appendix
\ref{sec:code} for code examples).
%
Finally, rather than varying the parameters of the existing equations
in the \texttt{FFW16} model, the equations themselves can be
fundamenally changed.  This requires defining a new physiological
model in a new C++ file, and compiling that into the package.  As an
example, in the \texttt{FFWdev} physiology, we have modified the
function determining allocation to reproduction.

With a physiological model in place, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an ordinary differential equation of
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:code} for more details). As the plant grows, allocation to
different tissue types varies and the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
costs and the growth rate of the plant.  This
shift in allocation by itself can generate the widely-observed hump-shaped pattern
of absolute height growth with size \citep[Fig. \ref{fig:plant}c;][]{King-2011} and the decline in relative mass growth rate with size from birth onwards \citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole-plant light compensation point (\textsc{wplcp}) can be computed (Fig.
\ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant
stops growing and is increasingly regarded as the most useful
measure of a species' shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
increases with plant size (Fig. \ref{fig:plant}d), due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
decreases with \textsc{lma} (Fig. \ref{fig:plant}d), because high \textsc{lma} species have slower leaf turnover \citep{Baltzer-2007, Lusk-2013}.

@@

Within patches of competing plants, competition for light generates
strong non-linear feedbacks on growth, survival and reproduction. In the
\texttt{FFW16} physiological model, we consider only the effect of
shading on rates of biomass production, although competition for other
resources such as nitrogen or water could also be considered by extending the
physiological model. We assume that patches are vertically but not horizontally
structured; in other words, we account for size differences but not spatial
layout within the patch. Similar assumptions are found in other \textsc{TSPMs}
\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
While ideally one would also consider spatial interactions within patches,
such models are very computationally demanding and also
impossible to solve in a deterministic fashion \citep{Shugart-1980,
Pacala-1996}. This assumption can be motivated from the observation
that competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

When modelling a patch of competing plants, the focus is on changes to the size-
density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density \(N\) of
individuals with traits \(x\) and height \(H\) within a patch of age \(a\).
Density is measured as individuals per unit height and per unit ground area.
Modelling the development of  \(N(x, H, a)\) requires that both the initial size
distribution and the inflow of new recruits be specified. We have primarily
been interested in dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b). A non-trivial extension could be added to allow some plants to survive disturbance \citep[following][]{Kohyama-1993}.

The {\plant} model offers two methods for modelling the dynamics of \(N(x,
H, a)\) in a competing population. In the first ``stochastic'' mode, users
specify an average seed rain and patch size, and {\plant} then generates a
vector of seed arrival events and simulates stochastic
development of the resulting finite-sized population.

For most applications, however, rather than modelling
dynamics within a finite-sized stochastic patch, it will be preferable to use {\plant}'s
``deterministic'' approximation. The deterministic mode assumes that
patches are very large, so that the population dynamics within the
patch approach their deterministic limit and can
be approximated via a Partial Differential Equation (\textsc{pde})
\citep{Deroos-1997, Kohyama-1993}.  The same \textsc{pde} is also thought to
capture the average behaviour across a large number of small patches
\citep{Moorcroft-2001}. Importantly, the ``deterministic'' mode is
much faster than the stochastic model and without the demographic noise
that arises in small populations (more details in Appendix \ref{sec:demography}).

Our approach for modelling deterministic size-structured population dynamics is
based on the Escalator Boxcar Train (\textsc{ebt}) technique
\citep{Deroos-1988, Deroos-1992, Deroos-1997, Brannstrom-2013}. The \textsc{ebt} solves
the \textsc{pde} describing development of \(N\) by approximating the
density function with a collection of cohorts spanning the size
spectrum. Following a disturbance, a series of cohorts are introduced
into each patch. These cohorts are then transported along the
characteristic equations of the \textsc{pde}. Biologically these are
the growth trajectories of individuals, conditioned on them surviving
(e.g. Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along the trajectories
(Fig. \ref{fig:patch}b).

We implement two extensions to the \textsc{ebt}: i) a new method for
estimating the size-density distribution and ii) a technique for
handling strong size-asymmetric competitive feedbacks, such as occur
under strong competition for light (see Appendix \ref{sec:demography}
for details). The original \textsc{ebt} proceeds by approximating the
first and second moments of the density distribution \(N\) within each
cohort \citep{Deroos-1997}.
%
In {\plant} we use a different approach. Instead of tracking first and
second moments of the size-distribution \(N\) within cohorts, we track a
point estimate of \(N\) along the characteristic for the boundary of
each cohort. We found this approach preferable to the approach of
approximating the first and second moments of \(N\) for each cohort
because it does not require us to maintain a separate set of equations for the smallest
cohort.

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population, an idea first applied in \citet{Falster-2011} and
described further in \citet{Falster-2015}.
In the original \textsc{ebt}, cohorts were introduced at a fixed time
interval. Under competition, however, the growth
trajectories of individuals born at similar times can diverge
substantially over time (Fig. \ref{fig:patch}a; Appendix
\ref{sec:demography} and \ref{sec:code}). This can become a problem later during patch
development because much of the population can be located between
widely-spaced cohorts, causing low numerical accuracy. To maintain
numerical accuracy, we use an iterative algorithm to adaptively refine
the cohort spacing until the trajectories of adjacent cohorts remain
adequately resolved throughout development of the patch (see Appendix
\ref{sec:demography} for details). This refinement gives rise to uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), with considerably tighter spacing of cohorts in
younger patches.

>>>>

\title{{\plant}: A package for modelling forest trait ecology \& evolution}
\author{Daniel S. Falster\textdagger\textasteriskcentered$^1$, Richard G. FitzJohn\textdagger$^1$, \\{\AA}ke Br{\"a}nnstr{\"o}m$^{2,3}$, Ulf Dieckmann$^3$, Mark Westoby$^1$}
\affiliation{
$^1$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
$^2$ Department of Mathematics and Mathematical Statistics, Ume{\aa} University, 901 87 Ume{\aa}, Sweden \\
$^3$ Evolution and Ecology Program, International Institute for Applied Systems Analysis, Schlossplatz 1, A-2361 Laxenburg, Austria \\
\textdagger These authors contributed equally.\\
\textasteriskcentered Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
A manuscript in consideration as an Applications Note for
publication in MEE as part of the Special Feature \emph{Demography
 beyond the Population}.\\
Word count: ~4200 words}
\date{}

@@

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers
\section{Summary}\label{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
 Population dynamics in forests are strongly size-structured:
 larger plants shade smaller plants while also expending
 proportionately more energy on building and maintaining woody stems.
 Although the importance of size
 structure for demography is widely recognised, many models
 either omit it entirely, or include only coarse approximations.
\item
 Here we introduce the {\plant} package, an
 extensible framework for modelling size- and trait-structured demography,
 ecology and evolution in simulated forests.
 At its core, {\plant} is an
 individual-based model where plant physiology and demography are mediated by
 traits. Individual plants from multiple species can be grown in isolation,
 in patches of competing plants, or in metapopulations under a
 disturbance regime. Patches can be modelled either as a stochastic
 realisation of a finite-sized population, or as their deterministic limit.
 These dynamics can be integrated into metapopulation--level
 estimates of
 invasion fitness and vegetation structure. Because fitness emerges as a
 function of traits, {\plant}
 provides a novel arena for exploring eco-evolutionary dynamics.
\item
 {\plant} is an open source R package and is available at
 \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
 Accessed from R, the core routines in {\plant} are written in C++.
 The package provides for alternative physiologies and for
 capturing trade-offs among parameters. A
 detailed test suite is provided to ensure correct behaviour of the code.
\item
 {\plant} provides a transparent platform for investigating how
 physiological rules and functional trade-offs interact with competition and
 disturbance regimes to influence vegetation demography, structure and
 diversity.
\end{enumerate}

@@

Plant growth and demography are fundamentally size- and
trait-structured, influencing dynamics over time-scales ranging from
instantaneous physiological effects to long-term evolutionary outcomes
\citep{Harper-1977, Westoby-2002}.
%
As an individual plant increases its leaf area, it increases its potential
to generate photosynthate.
%
On the other hand, as individuals grow larger, they must allocate increasing
fractions of their photosynthetic income to activities other
than building new leaves; for example, to maintaining support tissues
\citep{Givnish-1988, Enquist-2007} or to reproduction
\citep{Thomas-2011}.
%
Consequently, rates of growth, mortality and reproduction change with
individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
Ontogenetic
patterns of growth also vary with traits; for example,
leaf and wood construction costs influence growth rates
\citep{Wright-2010}, while seed size and height
at maturation determine start and end points of ontogenetic
trajectories \citep{Westoby-2002}.
%
Strong feedbacks emerge between individuals within a forest via
competition for light and other resources, such that the growth rate
of one individual depends on the size and traits of nearby
individuals \citep{Shugart-1980, Pacala-1996}.
%
Such feedbacks make it difficult to link observable species traits to
phenomena such as self-thinning, successional transitions, trait
evolution and species coexistence, without
modelling both individual growth and the competitive interactions between
individuals.

Although the importance of size structure for trait evolution, vegetation
dynamics and diversity has long been recognised \citep[e.g.,][]{Harper-1977,
 Shugart-1980, Huston-1987}, current research in this area is
dominated by models and theory that either omit size structure
entirely, or only include coarse approximations.
%
Theoretical investigations exploring questions about niche-based
species coexistence and ecological drift (neutrality) both rely
primarily on unstructured population models \citep[e.g.,][]{MacArthur-1967,
 Tilman-1985, Geritz-1998, Hubbell-2001, Calcagno-2006}.
%
It is common to assume competitive outcomes to be influenced by
size-related traits such as adult size and seed size, but detailed
size structure is rarely considered, at least in plant models
\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
%
Similarly, many of the models used to study global vegetation dynamics across
the last 20 years have not included size-structured demography
\citep[for comparisons of some major models, see][]{Sitch-2008,
Dekauwe-2014}.
%
While the importance of explicitly including size-structured dynamics is
increasingly recognised in both evolutionary ecology
\citep[e.g.,][]{Falster-2015} and vegetation dynamics
\citep[e.g.,][]{Moorcroft-2001, Purves-2008, Smith-2014,
Weng-2015, Sakschewski-2015}, our understanding of how these features
impact the ecology and evolution of vegetation communities remains
relatively limited.

In this note, we introduce the {\plant} package for R \citep{R-2015};
a framework for studying the effects of size structure and
trait variation on the demography of individual plants, of patches of
competing plants and of metapopulations structured by a
disturbance regime.
%
Our own purpose in developing {\plant} has been mainly to investigate how species
differing in traits may be able to coexist with one another
\citep[following ][]{Falster-2011, Falster-2015}; as size-structured models
provide unique opportunities for studying coexistence via differentiation in
successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
At the same time, we expect {\plant} to be useful for modelling other demographic and evolutionary phenomena.

Broadly, the {\plant} package falls into a class we refer to as trait-, size- and
patch-structured models \citep[\textsc{TSPMs}, following][]{Falster-2011}. \textsc{TSPMs} are direct
descendants from the ``gap'' models developed in the 1980s
\citep[e.g.,][]{Shugart-1980, Huston-1987, Kohyama-1993}; other modern examples
include \textsc{ED} \citep{Moorcroft-2001}, \textsc{LPJ- GUESS}
\citep{Smith-2014} and \textsc{LPJmL-FIT} \citep{Sakschewski-2015}. Common to
\textsc{TSPMs} is that they explicitly model size-structured competition
among individual plants within a metapopulation of ``patches'', where
individuals can differ in their traits and thus demographic behaviour.
\textsc {TSPMs} typically ignore the spatial structure of plants within
patches and the spatial structure of patches relative to one another. Instead,
they focus computational effort on resolving the dynamics of size-structured
competition for light and successional turnover. Differences among
\textsc{TSPMs} arise from the core physiological models used (describing
how the demography of individual plants responds to resource availability),
and also from the numerical techniques applied to scaling the physiological model
up to estimate emergent behaviours of individuals, patches and vegetation.

Below we describe the general approach of the {\plant} package, then provide
sections outlining potential applications at nested levels of ecological organisation.
More detailed technical documents are provided as supplementary information
(see Appendices; updated versions of these technical documents are
available within the {\plant} package itself). These documents outline \ref{sec:demography}) the system of
equations being solved when modelling the demography of plants, patches and metapopulations; \ref{sec:FF16}) a description of the core
physiological model used in {\plant}, and \ref{sec:code}) a compendium of worked code examples showing how to interact with the {\plant} package from R.

@@

The {\plant} package implements an individual-based model, meaning that the dynamics of the
population arise from rules specifying how individuals grow and interact.
Being driven by traits, the model can be extended to potentially very many species. The
core rules in {\plant} are about the short-term physiological
functioning of an individual plant and how this is influenced by its
traits, size and light environment. While {\plant} includes a core physiological model, the present paper uses it only as an example, to illustrate how one can take a given physiological model and simulate ecological and evolutionary outcomes. Thus, we do not aim to justify the particular physiological model used, to provide detailed
comparisons to data, or to promote this physiological model over those
in other \textsc{TSPMs}. (These are all topics deserving substantially
more attention than can be afforded here.)
An important feature of {\plant} considered as a software package is that it allows for users to substitute their own physiological models.

On top of the physiological model, {\plant} implements methods for
population dynamics and evolutionary dynamics (Fig. \ref{fig:schematic}), following methods
described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
phenomena can be studied at three levels: individual plants, stands of
competing plants and entire metapopulations. The dynamics at higher
levels of organisation arise as emergent properties, driven by growth
physiology, competition for light and disturbance (Fig. \ref{fig:schematic}).
{\plant} offers the capacity to model emergent phenomena in patches with
specified size, and also through a deterministic approximation
where specifying a patch size is not required. Trait evolution and community
assembly can then be modelled using the estimates of invasion fitness
provided by {\plant}.

\section{Effects of size, trait and light environment on the demography of individual plants}

@@

The default physiological model used in {\plant} largely reflects
that presented in \citet{Falster-2011, Falster-2015}, but with extensions
allowing for growth in plant diameters also to be estimated. We refer to this
default model as \texttt{FF16}, reflecting the initials of the first two authors
and publication year of this paper. The \texttt{FF16} model estimates rate
of biomass
production for a plant, given its size and the current light
environment. Gross photosynthetic income is calculated from the total
leaf area and the light distribution across the plant's canopy. Costs
of tissue respiration and turnover are subtracted. The remaining
biomass production is allocated between growth and reproduction. The key
outputs needed by higher levels of {\plant} are height
growth rate, mortality rate and rate of seed production (Fig. \ref{fig:schematic}). Additional quantities are also computed, such as total assimilation, as well as
respiration, turnover and allocation rates for different
tissues; these intermediates can also be accessed (for further details on the \texttt{FF16} model, see Appendices
\ref{sec:FF16} and \ref{sec:code}).

While we have implemented a particular physiological model
(\texttt{FF16}), the package is designed to allow arbitrary
additional physiologies to be introduced. Users are able to modify the
physiological model in two different ways. First, one can vary the
parameters of the \texttt{FF16} model directly (see Appendix
\ref{sec:FF16} for the list of parameters and Appendix \ref{sec:code} for code
examples). Second, we allow for changes in key parameters to bring
about changes in other parameters (a hyper-parameterisation), enabling the
straightforward modelling of trade-offs. For example, the
trait \textsc{lma} (leaf mass per unit leaf area) is used by default to estimate
the rate of leaf turnover ($k_l$), based on an observed scaling
relationship spanning across diverse vegetation and plant functional
types \citep{Wright-2004},
$$k_l = \beta_{kl1} \, \left(\phi / \phi_0\right)^{-\beta_{kl2}},$$
where $k_l$ and $\phi$ (\textsc{lma}) are both parameters of the core physiological
model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such
linkages are defined within a user-supplied hyper-parameterisation function
(see Appendix \ref{sec:FF16} for details, or Appendix
\ref{sec:code} for code examples).
%
Finally, rather than varying the parameters of the existing equations
in the \texttt{FF16} model, the equations themselves can be
fundamentally changed. This requires defining a new physiological
model in a new C++ file and compiling that into the package. As an
example, in the \texttt{FFBDWdev} physiology, we have shown how to modify the
function determining allocation to reproduction.

With a physiological model in place, {\plant} can be used to estimate
essential physiological rates for individual plants
(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
takes a given strategy and light environment and grows the plant,
producing a trajectory of plant size over time (Fig. \ref{fig:plant}a). This
is achieved by integrating an ordinary differential equation with a
size-dependent growth rate (see Appendices \ref{sec:demography} and
\ref{sec:code} for details). As the plant grows, allocation to
different tissue types varies; specifically, the composition of the plant
changes to include more stem and less leaf
(Fig. \ref{fig:plant}b). This, in turn, affects maintenance and turnover
costs and the growth rate of the plant. By itself, the
shift in allocation can generate the widely observed hump-shaped dependence
of absolute height growth on plant height \citep[Fig. \ref{fig:plant}c;][]{King-2011}, as well as the decline in relative mass growth rate with plant size from birth onwards \citep{Enquist-2007}.

By varying the light environment and measuring growth rate, the whole-plant light compensation point (\textsc{wplcp}) can be computed (Fig.
\ref{fig:plant}d). The \textsc{wplcp} is the light level where the plant
stops growing, increasingly regarded as the most useful
measure of a species' shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, the \textsc{wplcp}
increases with plant size (Fig. \ref{fig:plant}d), due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, the \textsc{wplcp}
decreases with \textsc{lma} (Fig. \ref{fig:plant}d), because high-\textsc{lma} species have slower leaf turnover \citep{Baltzer-2007, Lusk-2013}.

@@

Within patches of competing plants, competition for light generates
strong nonlinear feedbacks on growth, survival and reproduction. In the
\texttt{FF16} physiological model, we consider only the effects of
shading on rates of biomass production. Competition for other
resources such as nitrogen or water could be considered by extending the
physiological model. We assume that patches are vertically -- but not horizontally --
structured; in other words, we account for size differences, but not for the spatial
layout within patches. Similar assumptions are made in other \textsc{TSPMs}
\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
While one would ideally also consider spatial interactions within patches,
such models are very computationally demanding \citep{Shugart-1980,
Pacala-1996}. Simplifying spatial interactions within patches can be motivated from the observation
that competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

When modelling a patch of competing plants, the focus is on how the size-density-distribution \(N(x, H, a)\) changes with patch age \(a\), with the latter defined as the time that has passed since the last disturbance. This distribution describes the density \(N\) of
individuals with traits \(x\) and height \(H\) within a patch of age \(a\).
The density \(N\) is measured as individuals per unit height and per unit ground area.
Modelling the dynamics of \(N(x, H, a)\) requires that both the initial size-density-distribution and the inflow of new recruits be specified. We have primarily
been interested in the dynamics of a patch recovering from disturbance and have
therefore started with an empty patch and a constant flow of seeds from a global
seed rain (Fig. \ref{fig:schematic}b). A (non-trivial) extension would result from allowing some plants to survive disturbance \citep[following][]{Kohyama-1993}.

{\plant} offers two methods for modelling the dynamics of \(N(x,
H, a)\) in a competing population. In the first stochastic mode, users
specify an average seed rain and patch size, and {\plant} then generates a
vector of seed arrival events and simulates the stochastic
development of the resulting finite-sized population.

For most applications, however, rather than modelling
dynamics within a finite-sized stochastic patch, it will be preferable to use {\plant}'s
deterministic mode. This assumes that
patches are sufficiently large that population dynamics within a
patch approach their deterministic limit and can
be approximated via a physiologically structured population model \citep{Metz-1986, Deroos-1997, Kohyama-1993}. These models are often formulated as partial differential equations where boundary conditions and coefficients may depend on the population state. Such structured population models are also thought to
capture the average behaviour across a large number of small patches
\citep{Moorcroft-2001}. Importantly, the deterministic mode is
much faster than the stochastic mode, and eliminates the demographic noise
that inevitably arises in finite-sized populations (see Appendix \ref{sec:demography} for details).

Our approach for numerically solving deterministic size-structured population dynamics is
based on the characteristic method \citep{Angulo-2004}. This method describes the development of \(N\) by approximating the
density function along a collection of individual growth trajectories spanning the size
spectrum. Following a disturbance, a series of such trajectories are introduced
into each patch. These trajectories then change according to the growth
of the corresponding individuals, conditioned on their survival
(Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
to alter the density of individuals along these trajectories
(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train (\textsc{ebt}) method
\citep{Deroos-1988, Deroos-1997} used by \citet{Falster-2011} and
\citet{Falster-2015}, but not identical. An advantage of the characteristic method is that it allows direct approximation of the size-density-distribution. For the same reason, however, the characteristic method may be less suitable for resolving size-density-distributions in models where those distributions naturally develop sharp peaks (which happens, e.g., when plants completely cease to grow at a given height).

We implement a novel technique for
handling strongly size-asymmetric competitive feedbacks, such as those occurring
under strong competition for light (see Appendix \ref{sec:demography}
for details). This technique works for both the characteristic method and the Escalator Boxcar Train method. It involves an
adaptive refinement of the time points where new growth trajectories or cohorts are introduced
into the population, an idea first applied in \citet{Falster-2011} and
described further in \citet{Falster-2015}.
Normally, growth trajectories are introduced evenly, spaced out according to a fixed time
interval. Under size-asymmetric competition, however, the growth
trajectories of individuals born at nearby times can diverge
substantially over time (Fig. \ref{fig:patch}a; Appendix
\ref{sec:demography} and \ref{sec:code}). This can become a problem during patch
development, because much of the population can be located in between
widely-spaced trajectories, causing low numerical accuracy. To maintain
numerical accuracy, we use an iterative algorithm to adaptively refine
the cohort spacing until adjacent trajectories remain
adequately resolved throughout patch development (see Appendix
\ref{sec:demography} for details). This refinement gives rise to an uneven
distribution of cohort introduction times across patch age (Fig.
\ref{fig:patch}a), with a considerably tighter spacing of trajectories in
younger patches.
